---
title: "Europe Wildfire Season Tracker"
subtitle: "EFFIS rapid perimeters, updated weekly during fire season"
author: "Pierre Beaucoral"
date: today
execute:
  freeze: false
---

How bad is this fire season, and how would you actually know? This tracker
follows Europe's wildfire season as it unfolds. Satellites spot
large fires (roughly 30 hectares and up), and Europe's **EFFIS** service,
part of the Copernicus programme, draws the outline of each burned area it
can map. During the fire season (**June to September**) this site refreshes
**weekly**: a pipeline fetches the newest outlines, stores a dated copy under
`DATA/snapshots/`, and every chart re-renders from that copy. Fire sizes are
measured directly from the mapped shapes, in a map projection that keeps
areas honest, and each fire is assigned to the country it overlaps most. Same
method as the 2025 season analysis this project grew from.

```{r}
#| label: tracker-data

# Reads the SAME cached envelope object as posts/2026.qmd (identical cache
# key via build_envelope) -- nothing heavy is recomputed here.
library(sf)
library(dplyr)
library(tibble)
library(purrr)
library(lubridate)
library(ggplot2)
library(scales)
library(janitor)
library(rnaturalearth)

Sys.setenv(OGR_GEOJSON_MAX_OBJ_SIZE = "0")

source(file.path("R", "helpers.R"))
source(file.path("R", "geo.R"))
source(file.path("R", "cache.R"))
source(file.path("R", "theme.R"))
source(file.path("R", "pipeline.R"))
source(file.path("R", "plots.R"))
source(file.path("scripts", "latest_snapshot.R"))

snap <- latest_snapshot()
stopifnot(!is.na(snap))
eu <- get_eu()

envelope <- build_envelope(
  hist_years = 2017:2025, year_current = 2026L,
  snapshot_dir = snap, eu = eu,
  start_month = 6L, end_month = 9L
)

m <- envelope$meta
as_of_lab <- paste(day(m$as_of_date), month.name[month(m$as_of_date)])

cap_envelope <- sprintf(
  paste(
    "Cumulative burned area since 1 June: 2026 to date against the 2017–2025",
    "range (shaded band: min–max; dashed line: median). The 2026 line stops at",
    "its last mapped perimeter (%s)."
  ),
  as_of_lab
)
```

## Season at a glance, as of `r as_of_lab` `r m$year_current`

| How much has burned since 1 June | In how many separate fires | Compared with a typical recent season |
|:---:|:---:|:---:|
| **`r lab_si_ha(m$current_cum_ha)`** | **`r comma(m$current_n_fires)`** | **`r if (is.na(m$pct_vs_median)) "n/a (too early in season)" else sprintf("%.0f%%", m$pct_vs_median)`** |

: Totals count only the parts of each fire inside Europe (snapshot `r basename(snap)`). Fire *counts* are not comparable across years because the detection system changed; the median comparison uses burned *area*. Details on the [season page](posts/2026.qmd). {#tbl-glance}

```{r}
#| label: tracker-envelope
#| fig-width: 9.5
#| fig-height: 5.5
#| fig-cap: !expr cap_envelope

plot_envelope(envelope)
```

**The full picture** (hero and interactive maps, country rankings, raw and
normalized by land area, re-burn analysis, and what vegetation is burning)
lives on the **[2026 season page](posts/2026.qmd)**, updated with each
snapshot.
